Backpropagating Gravitational wave events

### 2022-05-20 src/ms.tex
Will works on intro and abstract.

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties; indeed,
measuring the variation of population synthesis parameter settings with binary
properties is one of the principal ways to extract information about binary
astrophysics from catalogs. Here we propose an inference pipeline to
simultaneously infer the posterior density of stellar progenitor parameters
(e.g. zero-age main sequence masses) and population synthesis parameter settings
controlling modeled binary evolution from gravitational wave observations.  Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO Virgo Kagra Gravitational Wave Transient Catalog (GWTC3).  We
showcase the effectiveness of this pipeline with a detailed study of the
progenitor properties and population synthesis settings that produce mergers
like the observed GW150914.
\end{abstract}

@@

\section{Introduction}
As the detection rate of gravitation waves (GWs) from merging
double-compact-object (DCO) binaries increases with the sensitivity of the
ground-based GW detector network
\citep{Abbott2018,LIGOScientificCollaboration2015,Buikema2020,Tse2019,Acernese2015,Acernese2019,Aso2013,Akutsu2021},
we are beginning to constrain the astrophysical processes which shape the
evolution of GW progenitor populations. One of the most common ways to study
progenitor populations of GW mergers is through population synthesis simulations
for a host of formation environments including isolated binary stars, triple
systems, globular clusters, nuclear star clusters, AGN disks, and more
\citep{Mandel2022}. \wf{Moar citations here} In these types of studies, the
population is forward modeled based on random samples from empirically or
theoretically motivated distributions of initial parameters which define the
age, metallicity, and intrinsic properties of the population like masses,
orbital separations, and eccentricities. These initial populations are then
evolved using a set of assumptions for how astrophysical processes impact the
evolution of the DCO merger progenitors, ultimately producing a synthetic
catalog of mergers which can be compared of observations from ground-based GW
detectors.

@@

While forward modelling entire populations has been the canonical pathway to compare simulations and GW observations,
it is not necessarily the most efficient way to constrain the underlying physical processes, or "hyperparameters", which drive the evolution of the population.
One of the most common skepticism of population synthesis studies is that these studies employ simulation sets which often 
cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
(e.g. common envelope efficiency) while holding all others fixed to some "default" values.
On top of this, these values are often not calibrated to data.
Even with recent developments in using emulators to speed up the simulation process, training the emulators 
still requires a significant amount of computation to cover a wide range of uncertainty.
As an example, to emulate a model with $10$ parameters, the simplest way to construct a training set of simulations is to 
run a grid which varies combinations of uncertain physics.
For 10 uncertain physical processes, if we consider $2$ variations for each physical process 
this would still require $1024$ simulations.
Because population synthesis simulations require hundreds to thousands of cpu hours,
training an emulator which spans a high dimensional uncertainty space remains an impossibility at present.

@@

This compounds with the fact that there is no guarantee the entire population should share the same fixed hyperparameters, 
as is often assumed in most population synthesis studies.
This suggests that there could be a large (possibly infinite) family of hyper-parameter distributions 
required to compare data to simulations which span the full set of uncertain physical models.
However, this issue is not a fundamental issue, but an apparent obstacle which arises from an inefficient approach to the problem.
Furthermore, each binary system could in principle come from a different formation environment and hence 
have different progenitor parameters as well as hyperparameters, which define uncertain physics associated with each environment.
In this case, it is advantageous to determine the posterior density in the progenitor and hyper-parameter joint space 
for each individual binary system, then perform population in analyses in that space.
\citet{Andrews2021} (hereafter A21) used \texttt{DartBoard} \citep{Andrews2018} 
to determine the ZAMS parameters which produce GW150914-like BBH merger based on posterior samples for GW150914 and a fixed 
set of hyperparameters which define assumptions for how isolated binary-star interactions proceed using 
\texttt{COSMIC}, a binary population synthesis code \citep{Breivik2020}.
However, the method was limited to a single event instead of the population.
More importantly, the method did not vary the hyperparameters while sampling the posterior density of the ZAMS parameters,
which limit its capability to be applied to understand a population of binary systems.

In this paper, we propose a method to "backward" model each GW event to its progenitor state 
while allowing the hyperparameters to vary across their full range of physical uncertainty
following a two-stage process.
First, we solve a root finding problem to obtain guesses that are likely to produce the desired system properties in the joint progenitor--hyper-parameter space.
We then sample the posterior in the joint space with Markov chain Monte Carlo (MCMC) algorithm that is initialized using the roots found in the previous stage.
The rest of the paper is structured as follows: We detail our method in Sec. ~\ref{sec:method}.
In Sec. ~\ref{sec:result}, we show the results of applying our method to a number of real events.
We discuss the implications of this work and future direction in Sec. ~\ref{sec:discussion}.

>>>>

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties; indeed,
measuring the variation of population synthesis parameter settings with binary
properties is one of the principal ways to extract information about binary
astrophysics from catalogs. Here we propose an inference pipeline to
simultaneously infer stellar progenitor parameters (e.g. zero-age main sequence
masses) and population synthesis parameter settings controlling modeled binary
evolution from individual gravitational wave observations of merging compact
binaries.  Our pipeline can efficiently explore the high-dimensional space of
population synthesis settings and progenitor system properties for each system
in a catalog of gravitational wave observations.  We apply our pipeline to
observations in the third LIGO Virgo Kagra Gravitational Wave Transient Catalog
(GWTC3).  We showcase the effectiveness of this pipeline with a detailed study
of the progenitor properties and population synthesis settings that produce
mergers like the observed GW150914.
\end{abstract}

@@

\section{Introduction}
As the detection rate of gravitation waves (GWs) from merging
double-compact-object (DCO) binaries increases with the sensitivity of the
ground-based GW detector network
\citep{Abbott2018,LIGOScientificCollaboration2015,Buikema2020,Tse2019,Acernese2015,Acernese2019,Aso2013,Akutsu2021},
we are beginning to constrain the astrophysical processes which shape the
evolution of GW progenitor populations. One of the most common ways to study
progenitor populations of GW mergers is through population synthesis simulations
for a host of formation environments including isolated binary stars, triple
systems, globular clusters, nuclear star clusters, AGN disks, and more
\citep{Mandel2022}. \wf{Moar citations here} Traditionally in these studies, the
population is studied through a Monte-Carlo approach.  Using theoretically
motivated distributions of initial parameters (age, metallicity, masses, orbital
separations, eccentricities, etc), initial populations are evolved with fixed
astrophysical assumptions (population synthesis parameter settings)to produce a
synthetic catalog of mergers which can be compared of observations from
ground-based GW detectors.

@@

While forward modelling entire populations has been the canonical pathway to
compare simulations and GW observations, it is not necessarily the most
efficient way to constrain the underlying physical processes, represented by
population synthesis ``hyperparameters,'' which drive the evolution of the
population.  \wf{I would not turn to ``skepticism'' here; instead, I would say
something like [but worded better] ``it is likely that the hyperparameter
settings that best represent the evolutionary physics vary across progenitor
parameter space in unknown ways, which makes the modeling very expensive if you
want to produce an entire population with one hyperparameter setting.
Additionally, comparing at the level of populations is not as efficient as
comparing event-by-event.  Ideally, we would like to \emph{hierarchically}
combine several observations to directly constrain the progenitor population
\emph{and} hyperparameter settings all at once from the observed catalog.''  I
see that these points come in the next paragraph, so one thing you could do is
dramatically shorten this paragraph and use it to lead into the next one.} One
of the most common skepticism of population synthesis studies is that these
studies employ simulation sets which often cover a small subdomain of the entire
parameter space of uncertain physics, or only vary one uncertain process per
simulation (e.g. common envelope efficiency) while holding all others fixed to
some "default" values. On top of this, these values are often not calibrated to
data. Even with recent developments in using emulators to speed up the
simulation process, training the emulators still requires a significant amount
of computation to cover a wide range of uncertainty. As an example, to emulate a
model with $10$ parameters, the simplest way to construct a training set of
simulations is to run a grid which varies combinations of uncertain physics. For
10 uncertain physical processes, if we consider $2$ variations for each physical
process this would still require $1024$ simulations. Because population
synthesis simulations require hundreds to thousands of cpu hours, training an
emulator which spans a high dimensional uncertainty space remains an
impossibility at present.

@@

This compounds with the fact that there is no guarantee the entire population
should share the same fixed hyperparameters, as is often assumed in most
population synthesis studies. This suggests that there could be a large
(possibly infinite) family of hyper-parameter distributions required to compare
data to simulations which span the full set of uncertain physical models.
However, this issue is not a fundamental issue, but an apparent obstacle which
arises from an inefficient approach to the problem. Furthermore, each binary
system could in principle come from a different formation environment and hence
have different progenitor parameters as well as hyperparameters, which define
uncertain physics associated with each environment. In this case, it is
advantageous to determine the posterior density in the progenitor and
hyper-parameter joint space for each individual binary system, then perform
population in analyses in that space. \citet{Andrews2021} (hereafter A21) used
\texttt{DartBoard} \citep{Andrews2018} to determine the ZAMS parameters which
produce GW150914-like BBH merger based on posterior samples for GW150914 and a
fixed set of hyperparameters which define assumptions for how isolated
binary-star interactions proceed using \texttt{COSMIC}, a binary population
synthesis code \citep{Breivik2020}. Crucially, the method did not vary the
hyperparameters while sampling the posterior density of the ZAMS parameters,
which limit its capability to be applied to understand a population of binary
systems.  \wf{This sounds sort of aggressive against DartBoard, and we probably
don't want to do that.  Instead, maybe re-focus the 'graph on
\emph{distinguishing} our method from DB; less DB sux and more ``here is how we
are not DB.''}

In this paper, we propose a method to "backward" model each GW event to its
progenitor state while allowing the hyperparameters to vary across their full
range of physical uncertainty following a two-stage process. First, we solve a
root finding problem to obtain guesses that are likely to produce the desired
system properties in the joint progenitor--hyper-parameter space. We then sample
the posterior in the joint space with Markov chain Monte Carlo (MCMC) algorithm
that is initialized using the roots found in the previous stage; the
initialization step makes the burnin and convergence of the MCMC stage
considerably faster than initializing with random parameter choices
\citep{Andrews2018,Andrews2021}. The rest of the paper is structured as follows:
We detail our method in Sec. ~\ref{sec:method}. In Sec. ~\ref{sec:result}, we
show the results of applying our method to a number of real events. We discuss
the implications of this work and future direction in Sec.
~\ref{sec:discussion}.
